% 00_abstract.tex — Abstract. Drafted by item P2-E8-1; restyled+condensed P2-E8; rewritten P2-R7-ABSTRACT;
% simplified 2026-06-24 (Maier: plain, precise, short sentences, non-native-readable; no literary register).
% Body only: this file is \input{} inside \abstract{...} in main.tex (no \section).
% NMI rule (document_check.md C): UNREFERENCED abstract — no \cite — and ~150-200 words.
Interpretability methods are judged by their faithfulness: whether they name the
true causes of a system's output. The field treats faithfulness as the test that
separates a real explanation from a merely plausible one. We test this on the
Understanding VCS Project, a bit-exact and differentiable copy of the Atari~2600 Video
Computer System (VCS).
The machine was never built to be interpretable. Yet it is fully known, so the
true cause of any output can be computed by exact intervention. We score thirty
interpretability methods for faithfulness against this known answer ($1$ = exact,
$0$ = chance), aggregated over 42 games.
Causal and intervention methods are more faithful than gradient and correlational
ones ($0.71$ versus $0.38$). The clearest case is a sprite's
position, where the naive gradient is provably zero: a causal method scores $1.000$
and standard saliency $0.115$. On this position regime the causal advantage is now
statistically established: the family-mean gap is $0.33$, with a 95 percent bootstrap
interval over the 42 games of $0.28$ to $0.38$ that excludes zero. But faithfulness is
not what is missing. When we switch on the port's
differentiable sampler, the naive gradient becomes non-zero---saliency rises
from $0.000$ to $0.791$ on Pong---yet its faithfulness stays low on most games and no
method recovers what a variable means.
Every method reports which signals matter, never what they mean. We name this
remaining distance the semantic gap, and we explore it in detail. Meaning is not in the
mechanism. For software it is fixed outside the code, in the documentation and the
data the program was written for. No method that reads only the mechanism can
reach it. The Understanding VCS Project offers a reusable benchmark with known
answers, for explainable AI, mechanistic interpretability, and software
verification.

% BIBTEX-NEEDED: (none) — the abstract is intentionally citation-free per the NMI
%   unreferenced-abstract rule (document_check.md C). No \cite keys introduced.
