% 01_intro.tex — Introduction (item P2-E8-2). Maier voice per ../STYLE.md.
% Arc: gap -> opportunity -> confirm -> the deeper move -> payoff.
% Central thread: faithfulness vs understanding / the semantic gap.
% Numbers cited from leaderboard.md / faithful_demo.md (committed §R) and figs 1,2.

\section{Introduction}\label{sec:intro}

When a method tells us what a system is doing, do we believe it because it is
\emph{true} or because it is \emph{plausible}? And once we have established that
its account is true, have we thereby \emph{understood} the program, or only named
its causes? These two questions sit at the centre of interpretability, and they
are not the same question. A neural network now drives a car, reads a scan, and
files a loan; before we trust such a decision we ask an explanation method why it
was made. The explanation is load-bearing for safety and for science alike, yet
on the systems we most want to explain we cannot check whether it is correct.
Saliency maps for an Atari agent look convincing and shift with the action, but a
convincing map is not a verified one~\cite{greydanus2018visualizing}. The field
has therefore converged on \emph{faithfulness} --- whether an explanation names
the system's \emph{true} causes --- as its quality criterion. Faithfulness is the
right criterion. The difficulty is that on a real neural network there is nothing
to measure it against.

This is the predicament Jonas and Kording made vivid by turning the
neuroscientist's toolkit on a microprocessor whose every transistor is
known~\cite{jonas2017could}. Lesions, tuning curves, and connectomics produced
rich, suggestive structure --- and none of it recovered how the chip actually
computes. Their lesson generalizes: when the ground truth is unknown, a method
that finds structure cannot be told apart from a method that finds the
\emph{right} structure. The same blind spot recurs across explainable AI, where
attribution maps are validated against each other or against human intuition
rather than against the system's mechanism~\cite{atrey2020exploratory}. Lacking a
reference, the community has quietly let faithfulness stand in for understanding,
and cannot test whether the proxy holds. We remove that excuse.

\paragraph{The opportunity.}
Paper~1 of this program built the missing reference: a bit-exact, differentiable
port of the Atari 2600 Video Computer System (VCS) that reproduces a real,
hand-authored 1977 machine to the byte and to the pixel~\cite{maier2025vcs,xitari}.
The VCS is software with a known specification, so the true cause of any output is
not a matter of opinion --- it is computable by exact intervention. We can clamp a
RAM cell, replace a register, occlude a ROM byte, re-run the machine
deterministically, and read off the change in the output. This gives us an
\emph{intervention oracle}: for every candidate cause of every output, the true
causal effect, measured rather than estimated. Figure~\ref{fig:platform} lays out
the platform and the oracle --- the three faces of the system we can observe and
intervene on (inputs, internal state, outputs) and the tiers of ground truth they
afford. On this machine, for the first time on a real and complex artifact, an
interpretability claim can simply be scored as right or wrong.

\paragraph{The move --- confirm.}
We turn the toolkit on this one system and score it. Across the neuroscience
battery, the attribution and explainable-AI methods, and the mechanistic-%
interpretability toolkit --- $31$ methods aggregated from $257$ per-game records
--- we place every method on two axes: faithfulness against the oracle, and a
documented plausibility proxy that captures how convincing the method's tradition
tends to look. The result confirms what the causal view predicts, now in numbers.
Causal and intervention methods are faithful; popular gradient and correlational
saliency methods are not. On discrete sprite-position outputs --- where the naive
gradient is provably zero, because position is set by strobe timing rather than by
a value the gradient can flow through --- the contrast is at its sharpest:
activation patching reaches faithfulness $1.000$, the oracle ceiling, while
vanilla saliency collapses to $0.000$, the naive-gradient floor, on every one of
the six core games. Aggregated over that regime, causal and intervention methods
average $0.412$ faithfulness against $0.068$ for the gradient and correlational
family, a gap of $0.344$. The popular method nonetheless carries the higher
plausibility ($0.9$ versus $0.5$): it looks more convincing precisely where it is
least correct. This is the field's danger zone --- conviction without correctness
--- and Fig.~\ref{fig:faithfulness} marks it out as a measured region of the
plane, not a cautionary metaphor.

\paragraph{The deeper move --- the boundary the leaderboard cannot cross.}
Securing faithfulness, however, is the smaller half of the problem. This is the
paper's central claim, and it is where a ground-truth machine lets us say
something the leaderboard alone cannot. With the same oracle we ask not only
\emph{which} methods are faithful but \emph{what a faithful method delivers}, and
the answer is a true causal-effect map or a true data-flow graph --- not the
variable's meaning and not the routine's algorithm. To make this precise we
require that an explanation be right only if it is \textbf{F}aithful (its claimed
causes are the true causes), \textbf{S}ufficient (it regenerates the behavior
under held-out interventions), and \textbf{M}inimal at the algorithmic level
(no spurious parts, at the granularity of the program's variables and circuits):
$\mathrm{right} = \mathbf{F}\wedge\mathbf{S}\wedge\mathbf{M}$ (Section~\ref{sec:triad}).
The committed data then show the gap directly, and in both directions at once. An
automatically discovered circuit for Breakout is graph-correct and minimal yet
does not reproduce the behavior it explains ($\mathbf{F}=1.0$, $\mathbf{M}=1.0$,
$\mathbf{S}=0.44$): a correct wiring diagram is not the computed function. A
learned feature dictionary matches a named game variable perfectly (matched
fraction $1.0$) while being causally almost inert ($\mathbf{F}=0.04$): looking
like a known variable is not being how the program computes it. And a concept can
be linearly decodable from the state yet provably unused by the computation --- the
familiar trap that what is \emph{present} need not be \emph{used}. The residual
between a faithful attribution and an understanding of the computation is therefore
not philosophy. On this machine it is a number. We name it the \emph{semantic
gap}, and making it a measured quantity is the contribution we care about most.

\paragraph{Why a 1977 chip judges methods built for neural networks.}
The obvious objection is that deterministic, hand-coded software is the wrong place
to judge methods designed for learned, distributed, stochastic networks. Our
answer, stated here and not deferred to the discussion, is that the VCS is a
\emph{necessary-condition screen}, not a predictor of success. A method that fails
\emph{here} --- with perfect ground truth, full observability, exact
interventions, and no learning to obscure the mechanism --- has no business being
trusted on a neural network, where every one of those advantages is lost. Passing
here is necessary, not sufficient; failing here is decisive. The same logic
strengthens the harder claim: if even the strongest causal account a fully
observable machine permits stops short of meaning, then on a network --- less
observable, without clean ground truth --- faithfulness is \emph{a fortiori} not
understanding. The gap we measure here is a floor on the gap there. The VCS also
exhibits the very properties that make interpretation hard --- RAM cells aliased
and reused across routines, race-the-beam timing that distributes a computation
across a scanline, and registers whose meaning depends on context. What it does
not capture --- learned features, scale, stochasticity --- we state plainly and
leave to the program's later papers on learned agents.

\paragraph{An honesty contract.}
Because the claim is sharp, we are deliberate about its limits. We do \emph{not}
claim that any method recovered the program's meaning, algorithm, variables, or
design from scratch. Every semantic label we use is imported from external sources
--- the AtariARI and OCAtari annotation
sets~\cite{anand2019unsupervised,delfosse2023ocatari} --- and then only
\emph{causally verified} on our substrate, never \emph{discovered} by a method
under test. When an alignment method reports perfect accuracy against such a label,
it is right by construction, because it is checking a variable we supplied. Probing
shows that information is present, not that it is used~\cite{hewitt2019designing};
our intervention oracle is the strictly stronger test. Accordingly, Paper~2
\emph{defines and measures} the semantic gap; \emph{closing} it --- recovering a
system's documentation and design from nothing but the system --- is the companion
work, not this paper. We measure the gap; we do not pretend to have crossed it.

\paragraph{In this Article.}
We make four contributions. \emph{First}, a ground-truth interpretability benchmark
on a single fully known, differentiable, and \emph{real} complex system, with the
necessary-condition-screen framing for reading a score on it; the headline is that
faithful equals causal, that popular saliency is plausible-but-unfaithful and
provably zero on discrete position outputs, and that the faithfulness-versus-%
plausibility danger zone is quantified (Fig.~\ref{fig:faithfulness}). \emph{Second},
and most importantly, the semantic gap made measurable: an operational definition,
$\mathrm{right}=\mathbf{F}\wedge\mathbf{S}\wedge\mathbf{M}$, under which even a
perfectly faithful method is shown on committed data to recover the causal wiring
and not the semantics --- graph-correct circuits that fail sufficiency, perfectly
matched features that are causally inert, decodable-but-unused concepts.
\emph{Third}, a quantitative re-run of the Jonas--Kording experiment, scoring the
neuroscience battery against a known mechanism and supplying the high-faithfulness
causal contrast they could not. \emph{Fourth}, the first ground-truth calibration
of mechanistic interpretability on a real artifact rather than a synthetic or
compiled one, together with the demonstration that circuit recovery yields a graph,
not the computed function. Read as a whole, the study offers a reusable benchmark
for explainable AI, mechanistic interpretability, AI safety, and software
verification alike, and reframes the field's open problem. The hard part of
interpretability is not faithfulness. It is the semantic gap that remains once
faithfulness has been secured --- and we believe that naming and measuring it is
the necessary first step toward closing it.


% BIBTEX-NEEDED: keys cited in this section that are NOT yet in references.bib at draft time
% are: maier2025vcs, anand2019unsupervised, hewitt2019designing (please MERGE these three).
% Already present (do NOT duplicate): jonas2017could, atrey2020exploratory,
% greydanus2018visualizing, xitari, delfosse2023ocatari. Full entries below for all keys
% I cite, for completeness; SM should de-duplicate against references.bib before merging.
%
% @article{jonas2017could,
%   title   = {Could a Neuroscientist Understand a Microprocessor?},
%   author  = {Jonas, Eric and Kording, Konrad Paul},
%   journal = {PLoS Computational Biology},
%   volume  = {13}, number = {1}, pages = {e1005268}, year = {2017},
%   publisher = {Public Library of Science}, doi = {10.1371/journal.pcbi.1005268}}
%
% @inproceedings{atrey2020exploratory,
%   title     = {Exploratory Not Explanatory: Counterfactual Analysis of Saliency Maps for Deep Reinforcement Learning},
%   author    = {Atrey, Akanksha and Clary, Kaleigh and Jensen, David},
%   booktitle = {International Conference on Learning Representations (ICLR)},
%   year      = {2020}}
%
% @inproceedings{greydanus2018visualizing,
%   title     = {Visualizing and Understanding Atari Agents},
%   author    = {Greydanus, Sam and Koul, Anurag and Dodge, Jonathan and Fern, Alan},
%   booktitle = {International Conference on Machine Learning (ICML)},
%   pages     = {1792--1801}, year = {2018}, organization = {PMLR}}
%
% @inproceedings{anand2019unsupervised,
%   title     = {Unsupervised State Representation Learning in Atari},
%   author    = {Anand, Ankesh and Racah, Evan and Ozair, Sherjil and Bengio, Yoshua and C\^{o}t\'{e}, Marc-Alexandre and Hjelm, R Devon},
%   booktitle = {Advances in Neural Information Processing Systems (NeurIPS)},
%   volume    = {32}, year = {2019}}
%
% @inproceedings{delfosse2023ocatari,
%   title     = {OCAtari: Object-Centric Atari 2600 Reinforcement Learning Environments},
%   author    = {Delfosse, Quentin and Bl{\"u}ml, Jannis and Gregori, Bjarne and Sztwiertnia, Sebastian and Kersting, Kristian},
%   booktitle = {Reinforcement Learning Conference (RLC)}, year = {2024},
%   note      = {arXiv:2306.08649}}
%
% @inproceedings{hewitt2019designing,
%   title     = {Designing and Interpreting Probes with Control Tasks},
%   author    = {Hewitt, John and Liang, Percy},
%   booktitle = {Conference on Empirical Methods in Natural Language Processing (EMNLP)},
%   pages     = {2733--2743}, year = {2019}}
%
% @misc{maier2025vcs,
%   title  = {A Bit-Exact, Differentiable Atari 2600 (Paper 1)},
%   author = {Maier, Andreas and Bayer, Stefan and Krauss, Patrick},
%   year   = {2025}, note = {companion paper; this program}}
