% 03_methods.tex — Methods (emulator, oracle, F-and-S-and-M triad, metrics).
% Drafted by item P2-E8-4. Voice: ../STYLE.md (Andreas Maier). Numbers: committed
% leaderboard/faithful_demo + experiment_design.md §0–§3. Figure: \ref{fig:platform}.

\section{Methods}\label{sec:methods}

Our method rests on a single idea. We study an interpretability question on a machine whose
answer we can compute exactly. We describe in turn the substrate that makes the ground truth
obtainable, the oracle that delivers it, the tiered semantics we attach to it, and the
correctness triad that scores understanding as numbers. The per-method
setups live in the Supplementary Information. Fig.~\ref{fig:platform} sketches the pipeline.

\subsection{The differentiable, bit-exact VCS substrate}\label{sec:substrate}

The subject of this study is the Atari Video Computer System (VCS) itself, not a learned
agent. It is a real, complex artifact never designed to be interpretable. Yet we can
measure its every causal dependency exactly. Paper~1 built it, a bit-exact, differentiable
port of the 1977 VCS. The port is validated bit-for-bit against the \texttt{xitari} C++
reference on all 64 games of the Paper-1 roster, in random-access memory and rendered screen,
within the validated conformance window~\cite{maier2025vcs,xitari}. We use it in two
interchangeable, cross-checked realizations, \texttt{jutari} (Julia) and \texttt{jaxtari}
(JAX/Python).

The substrate grants four capabilities that no neural network offers for free. The first is
an exact forward map,
\begin{equation}
  y \;=\; \mathrm{VCS}\big(\mathrm{ROM},\, \mathrm{inputs}[0{:}t],\, s_0\big),
  \label{eq:vcs}
\end{equation}
that is, the output $y$ at step $t$, a pixel, a score, or a game event, is a deterministic
function of the cartridge bytes, the inputs up to $t$, and the initial state $s_0$. Re-running
reproduces $y$ to the bit. The second is exact interventions. We set any cause
$u$, a ROM byte, a RAM cell, a CPU/TIA/RIOT register, an opcode (a CPU instruction byte), or a joystick input, to any
value and re-run Eq.~\eqref{eq:vcs} deterministically. This realizes the causal operator
$\mathrm{do}(u)$ on a real machine rather than an assumed world model. The third is full
state observability. We record every bit of RAM and every register along the trajectory, so
the ``activations'' we interpret are the program's own variables. The fourth is a
content-path gradient $\partial y / \partial u$ through a straight-through estimator (STE,
a surrogate gradient whose forward pass still returns the exact discrete value)
formulation whose forward pass is bit-exact to the hard machine at any finite
temperature (a smoothing parameter)~\cite{bengio2013estimating,jang2017gumbel}, batched on the GPU. A neural
interpretability study works with three things: the inputs, the state, and the outputs. Our
substrate gives us all three.

We respect one limit throughout. Paper-1 bit-exactness holds within a bounded frame window
on a fixed action stream, on the order of 30 frames of RAM and 60 frames of screen. Every
scored claim is taken inside that validated horizon. Long-horizon end-to-end gradients
are not needed for our single-step metrics, so we do not assume them.

\subsection{The ground-truth oracle}\label{sec:oracle}

The oracle is central to the method. For any output $y$ and cause $u$, it returns the true
effect of $u$ on $y$. We score every method against it. Its primary construction is the exact
intervention oracle: we intervene on $u$ by occlusion, clamping, replacement, or resampling,
re-run the machine deterministically, and record the change in the output,
\begin{equation}
  \Delta y(u) \;=\; \mathrm{VCS}\big(\dots,\, \mathrm{do}(u{:=}v'),\, \dots\big) \;-\; y .
  \label{eq:doy}
\end{equation}
The true causal effect of a cause is the difference its edit makes to the output. We measure
it by running the altered machine. We assume no world model and no smoothness. The collection
$\{\Delta y(u)\}$ over all causes is the true causal map of $y$, the reference for the
faithfulness axis of every method.

The output $y$ comes in five regimes. The regime fixes what a method must predict and whether
a gradient is even meaningful. A \emph{content} output is a register, color, or graphics-bit
value that flows smoothly into the pixel, where the naive gradient is valid. A
\emph{position/index} output is a sprite, missile, or ball coordinate set by strobe timing.
Here the naive gradient vanishes by Prop.~\ref{prop:zero}, and only the bilinear-sampler
surrogate or the intervention oracle applies. An \emph{event} output is a discrete state
change, a collision latch (a one-bit hardware flag) or a life lost, also gradient-zero. A \emph{score} output is the
game's running counter, valid for the gradient only where it varies continuously. A
\emph{pixel} output is the rendered framebuffer (the array of screen pixels) value. It is content-valid where it tracks a
smooth cause and index-zero where it tracks a position. The Supplement lists, per game, which
outputs each method is scored on and the per-output gradient verdict.

The oracle uses several kinds of intervention. We list each one, because the kind limits what
a score means. Occlusion and clamping are surgical but \emph{off-manifold}: they can set a
byte to a value the running program would never write. Replacement from a donor trajectory is
\emph{on-manifold}: the substituted value is one the machine actually produced elsewhere.
Resampling draws the cause from its own marginal over the trajectory. Counterfactual edits are
valid under the emulator by construction, since any byte assignment yields a well-defined
deterministic re-run. Off-manifold interventions are still valid for oracle scoring. The
oracle measures a causal effect, not a likelihood. We ask what the output would be if $u$ were
$v'$. The bit-exact machine answers this exactly for any $v'$, on- or off-manifold. Where a
method's own claim is about typical behavior rather than causal dependence we fall back to the
on-manifold donor and resampling variants and flag the choice in the Supplement.

The content-path gradient $\partial y / \partial u$ is the second way to build the oracle. We
read it through the differentiable substrate. It is fast and dense, and it agrees with the
intervention oracle on the content path, the register, color, and graphics-bit values that
flow smoothly into the pixel. Yet, one important limit applies to the whole study.
The straight-through gradient routes to the content path only. So a discrete index, a
sprite's screen position the VCS sets by the timing of a strobe write (a write that latches the sprite's horizontal position), has a naive gradient
that vanishes under the hard forward map. We make this precise.

\begin{proposition}[Zero naive gradient for discrete-index outputs]\label{prop:zero}
Let an output $y$ depend on a cause $u$ only through a discrete index
$n(u)=\lfloor g(u)\rfloor\in\mathbb{Z}$, the integer count of color clocks (each one pixel-time of the scanning beam) between a strobe
write and the position-counter reset that the VCS uses to place a sprite, missile, or ball,
with $g$ continuous, piecewise linear, and non-constant (strictly monotone on each linear
piece). Under the hard (bit-exact) forward map the rendered output is $y=h(n(u))$ for a fixed
render function $h$. Then $n(\cdot)$ is locally constant wherever $g(u)\notin\mathbb{Z}$, so
$\partial y/\partial u=0$ at every such point, and $y$ is discontinuous on the set
$\{u:g(u)\in\mathbb{Z}\}$, which is measure zero precisely because $g$ is strictly monotone on
each piece, and there the gradient is undefined. Hence the naive gradient is zero wherever the
derivative exists: a tiny change in $u$ cannot move the sprite by part of a pixel. Any method
that reduces to this naive gradient inherits the failure on position, index, and event outputs.
\end{proposition}

The proposition states only the hard-map limit; it does not say that no useful gradient
exists. Two surrogate constructions of Paper~1~\cite{maier2025vcs} are explicit exceptions and
are \emph{not} covered by it: the straight-through estimator, whose forward pass stays
bit-exact to the hard machine at any finite temperature while routing a non-zero
gradient along the content path~\cite{bengio2013estimating,jang2017gumbel}, and the bilinear
sampler that interpolates the index boundary to recover a usable position
gradient~\cite{jaderberg2015spatial}. The companion paper constructs both and validates the
bit-exactness this proposition assumes. We import that construction and prove the vanishing
inline, so the present paper stands alone. We therefore avoid any blanket claim that
``gradients are provably zero.'' We use instead the scoped claim of Prop.~\ref{prop:zero}: the
\emph{naive} gradient for discrete position/index outputs is zero under the hard forward map,
and methods that reduce to it inherit that failure. Surrogate and bilinear-sampler gradients
are exceptions.

We make the faithfulness audit as strict as possible. The position regime leaves one
objection open, and we close it. A reader might object that the naive zero of
Prop.~\ref{prop:zero} is just a fixable artifact. So we fix it: we turn on the Paper-1
bilinear sampler. The sampler interpolates the index boundary of Prop.~\ref{prop:zero} in the
spatial-transformer manner. This makes the discrete position output a continuous function of
the position RAM byte rather than a step. So the gradient methods receive a non-zero surrogate
position gradient where the hard map gave them nothing~\cite{jaderberg2015spatial}. We score
both conditions, the naive forward and the sampler-on forward, against the same intervention
oracle of Eq.~\eqref{eq:doy}. For each candidate cause the per-cause attribution is
correlated, by the raw Pearson $\rho$ of \Sectionname\ref{sec:triad}, against the oracle's true
$|\Delta\,\text{position-pixel}|$. So the two conditions differ only in the surrogate the
method is handed, not in how it is judged.
Alongside faithfulness we record a second, deliberately stricter property. The
\emph{semantic recovery} of a method is a yes/no question: does the method, on its own, emit a
game-concept name, the word ``ball'' or ``missile'' attached to the byte it points at,
rather than a cause that an external label only names later? An attribution map returns weights
over causes and never a name, so for it the answer is no by construction: it recovers no meaning.
Faithfulness and semantic recovery are independent, and they are of different kinds. Faithfulness
is a graded score for whether the method names the right cause. Semantic recovery is a yes/no
property: whether it names any meaning at all. The keystone experiment of
\Sectionname\ref{sec:results_compare} reports both for the naive and sampler-on conditions on the
position regime.

Hence, for content outputs the gradient is a legitimate companion oracle, while for position,
index, and event outputs the intervention oracle of Eq.~\eqref{eq:doy} is the sole ground
truth and every gradient-based method is scored as a method under test. This position regime
is where the sharpest separations in our results appear. Where the two overlap we report their
correlation and flag, rather than hide, the points at which they disagree.

\subsection{Tiered ground truth and how we obtain it}\label{sec:tiers}

The ground truth comes in three tiers. The tiers are not just bookkeeping; they support the
paper's central argument. Tier~1 (T1) is the exact causal map of Eq.~\eqref{eq:doy},
derived internally from the substrate and available for all 64 games. Tier~2 (T2) is the
hardware roles and the read/write data-flow graph, also exact and internal across all 64
games. T1 and T2 describe what the machine does internally, and need no external label.

Tier~3 (T3) is different in kind. This difference matters for our argument. It attaches a
game-concept name to a cell or feature, ``the ball's $x$ position'' or ``the score.'' These
names cannot be computed from the substrate alone. They are facts about what the program
\emph{means}, which is separate from what it \emph{computes}. We import T3 from two public
sources and make it causal. AtariARI provides labels for 22 games, read from commented
disassemblies, with raw RAM coordinates not aligned to the rendered
position~\cite{anand2019unsupervised}; OCAtari covers more than 40 games and, in its RAM
mode, applies the render-time offsets that fix that misalignment~\cite{delfosse2023ocatari}.
We import both as candidates and verify each causally by setting the byte, re-rendering, and
confirming on the bit-exact framebuffer that the named object moves to the predicted place.
This upgrades a correlational label to a verified-causal one and auto-corrects the offsets;
further labels we discover by RAM-to-framebuffer correlation and intervention. The headline
core set is six games, Pong, Breakout, Space Invaders, Seaquest, Ms.\ Pac-Man, and Q*bert,
each carrying both sources, all in the Paper-1 bit-exact roster. Together they span
paddle-and-ball, shooter-with-shields, resource-meter, multi-agent-pursuit, and discrete-grid
logic. The remaining games form a breadth set scored on T1 and T2 only.

We are careful about what this gives us. All T3 labels are external and only verified,
never discovered from scratch. So a method that matches a supplied concept has aligned to a
given meaning, not recovered it. This is exactly why a probe can show that information is
\emph{present} without showing it is \emph{used}~\cite{hewitt2019designing}. Our intervention
test asks whether the byte \emph{causes} the named object to move, so it is strictly stronger
than decoding it. Recovering the program's documentation and design from the bare machine is a
separate problem, the subject of the companion paper.

\subsection{The F\,$\wedge$\,S\,$\wedge$\,M correctness triad}\label{sec:triad}

A faithful attribution is not yet an understanding. To measure understanding, we need an
operational criterion. We adopt one in the spirit of Marr's levels of
analysis~\cite{marr1982vision} and of Jonas and Kording's warning that finding structure is
not the same as explaining it~\cite{jonas2017could}: each axis is first a continuous score,
and the boolean ``right'' is a thresholded reading of those scores. We score an
explanation $\hat{E}$ of how an output $y$ arises by three numbers, faithfulness
$\mathrm{F}_{\mathrm{score}}(\hat{E})$ and minimality $\mathrm{M}_{\mathrm{score}}(\hat{E})$
in $[0,1]$ and the held-out sufficiency $\mathrm{S}_{\mathrm{score}}(\hat{E})$ in $[-1,1]$,
and call it \emph{right} when all three clear their thresholds,
\begin{equation}
  \mathrm{right}(\hat{E}) \;=\; \mathbf{1}\!\left[\,
  \mathrm{F}_{\mathrm{score}}\!\ge\!\tau_{\mathrm{F}} \,\wedge\,
  \mathrm{S}_{\mathrm{score}}\!\ge\!\tau_{\mathrm{S}} \,\wedge\,
  \mathrm{M}_{\mathrm{score}}\!\ge\!\tau_{\mathrm{M}}\,\right],
  \label{eq:triad}
\end{equation}
with $\tau_{\mathrm{F}}=\tau_{\mathrm{S}}=\tau_{\mathrm{M}}=0.5$ throughout. Results report the
continuous scores; the predicate is a derived summary, not the primary quantity. All three are
defined against the oracle, not against any method under test, which keeps the criterion
non-circular.

Faithfulness asks whether the claimed causes are the \emph{true} causes, scored against the
intervention oracle of Eq.~\eqref{eq:doy}. The per-record metric depends on the explanation's
output type, so $\mathrm{F}_{\mathrm{score}}$ is a single number assembled from three
type-matched primitives. The committed score is the bare effect-agreement primitive, not a
rescaled one. For a real-valued attribution map we report the Pearson (or Spearman, for ranked
outputs) correlation $\rho(\hat{E},\{\Delta y(u)\})$ between the attribution and the true causal
effects, scored \emph{raw}. A method whose attribution carries no causal signal therefore lands
at $\rho=0$, not at the $0.5$ of an affine $\tfrac{1}{2}(1+\rho)$ remap. This is what makes
the position regime informative. Under Prop.~\ref{prop:zero} the naive gradient is identically
zero there, and its correlation with the oracle is $0$. The reported faithfulness is therefore
$0.00$ and not a $0.5$ floor. The two committed estimators differ only in how the raw
correlation is summarized across the scored games. Phase~B reports the raw correlation
directly (e.g.\ occlusion at $F=0.687$), so a negative correlation is recorded as a negative
number. In Phase~A's single-unit and tuning analyses the per-game correlations can be negative,
so we clip each per-game correlation at $0$ and then average across games (e.g.\ the tuning
analysis A3 at $0.213$, the clip-at-zero mean of per-game correlations
that include negatives). For the same attribution map we also report, as quality measures,
deletion/insertion area-under-curve (deleting or inserting the top-attributed inputs and
tracking the change in output) and precision at $k$ against the true causal top-$k$,
$\mathrm{P@}k=|\,\mathrm{top}_k(\hat{E})\cap\mathrm{top}_k(\{\Delta y(u)\})\,|/k$. For a
graph-valued explanation (Phase~A connectomics, Phase~C circuit discovery) we use the edge
$F_1$ against the true data-flow graph, the harmonic mean of precision and recall over recovered
edges. For Phase~C patching and mediation we use the causal-effect-agreement correlation against
the true $\{\Delta y(u)\}$. The final $\mathrm{F}_{\mathrm{score}}$ is the per-record primary
metric averaged first within a game over its output records, then with equal weight across the
scored games, so no game-rich method dominates; confidence intervals are 95\% bootstrap over
games (\Sectionname\ref{sec:scoring}). The three Phases differ only in which primitive is primary: Phase~A
uses graph $F_1$ for the data-flow methods and the clip-at-zero mean correlation for the
single-unit and tuning analyses, Phase~B uses the raw effect-agreement correlation $\rho$ with
deletion/insertion AUC and $\mathrm{P@}k$ as the attribution map's quality measures, and Phase~C
uses the causal-effect-agreement correlation for patching and mediation and graph $F_1$ for
circuit discovery. The leaderboard score is therefore a single scalar per method, but it is
multi-metric in origin, and the per-method choice is recorded in the Supplement.

Sufficiency asks whether the explanation could regenerate the behavior under interventions it
has not seen. We treat $\mathrm{S}_{\mathrm{score}}$ as a held-out predictive estimator: the
explanation's per-cause claim, its attribution or the effect read off its causal sites, is
read on a calibration split of interventions and used to predict the true effect on a disjoint
held-out split of $\mathrm{do}(u)$ edits, each compared to the bit-exact re-run. We score it by
the held-out predictive \emph{correlation} between the claimed and the true $\Delta y$ on that
held-out split (Pearson, or Spearman for ranked outputs), so
$\mathrm{S}_{\mathrm{score}}\in[-1,1]$: a value near zero means the explanation predicts no
better than the unperturbed output, and a negative value means it predicts \emph{worse}, its
held-out claim anti-correlated with the truth (the sparse autoencoder reaches
$\mathrm{S}=-0.46$ on Pong, \Sectionname\ref{sec:results-c}). Where the output admits only an
exact match, a discrete index or event, we report instead the fraction of held-out
interventions predicted correctly within a fixed tolerance, the $[0,1]$ special case of the
same held-out test. We deliberately do not define it through interchange interventions or
scrubbing. Those are themselves mechanistic-interpretability methods under test in Phase~C, so
using them would make the criterion circular. All three phases report $\mathrm{S}$ where the
calibration and held-out split is defined, and its bootstrap intervals follow
\Sectionname\ref{sec:scoring}.

Minimality asks whether the explanation is parsimonious and lives at the algorithmic level of
the machine's own module decomposition. We define
$\mathrm{M}_{\mathrm{score}}=|U^\star|/|U_{\hat{E}}|\in(0,1]$, the ratio of the size of the
true minimal cause set $U^\star$, the smallest set of registers, RAM cells, opcodes, or
program variables whose intervention reproduces $y$, to the size $|U_{\hat{E}}|$ of the cause
set the explanation actually names. An explanation that names exactly the minimal set scores
$1$; one that records the whole machine state pays for every spurious cell. The target level
is the algorithmic unit, the register, RAM cell, opcode, module, or program variable, not the
pixel. Whole-state recording names all $128$ RAM cells, where the true cause set averages a
handful. It therefore scores $\mathrm{M}_{\mathrm{score}}=0.08$ across the scored games. Its
faithfulness is perfect because it withholds nothing, but it is maximally non-minimal.

Two reporting axes carry the headline comparison, faithfulness against the ground truth and a
plausibility proxy. The plausibility proxy is a documented, tradition-level value, not a
human-subjects measurement, and we report it as such. It is assigned per method tradition, not
per method instance, from the published evidence on how convincing each tradition's
explanations are taken to be, before any faithfulness score on our machine is computed, so the
proxy cannot be tuned to the result. The per-tradition assignment, its rubric, and its
sources are tabulated in the Supplement (\Sectionname\ref{supp:plausibility}). There a sensitivity
analysis perturbs the proxy under additive, rank-preserving, rank-jittering, and
leave-one-tradition-out schemes. It finds the faithfulness-versus-proxy rank correlation
stays negative throughout (from $-0.79$ to $-0.07$), so the headline contrast survives every
alternative proxy assignment we tried. The proxy lets us find explanations with high
plausibility but low faithfulness: explanations that convince but are wrong. F is satisfied by
the wiring; S and M must be satisfied by the \emph{computation}. Separating them lets us
measure, later, what faithfulness alone leaves out.

\subsection{Methods under test, games, and aggregation}\label{sec:scoring}

We apply three traditions to the machine and score each by the triad of
Eq.~\eqref{eq:triad}. The neuroscience battery (Phase~A) runs connectomics and data-flow
recovery, single-unit lesions, tuning curves, pairwise and global correlation structure,
pooled-activity spectra, Granger causality, dimensionality reduction, and whole-state
recording, scoring each against the known mechanism to quantify Kording's lesson at the
low-faithfulness baseline~\cite{jonas2017could}. The attribution and XAI
battery (Phase~B) runs gradient saliency and its variants, integrated gradients, occlusion,
perturbation masks, randomized masking, LIME, and Shapley-style
attribution~\cite{simonyan2014deep,sundararajan2017axiomatic,zeiler2014visualizing,fong2017interpretable,petsiuk2018rise,ribeiro2016why,lundberg2017unified};
Grad-CAM and its variants and attention
rollout~\cite{selvaraju2017gradcam,chattopadhyay2018gradcampp,abnar2020quantifying} require
neural layers or a learned policy and are recorded as not applicable. The
mechanistic-interpretability battery (Phase~C) runs activation patching and causal
mediation, interchange interventions and distributed alignment search, attribution and path
patching, automatic circuit discovery, sparse autoencoders, causal scrubbing, linear
probing with control tasks, and the logit and tuned
lens~\cite{vig2020investigating,meng2022locating,geiger2021causal,geiger2023finding,nanda2023attribution,wang2022interpretability,conmy2023towards,cunningham2023sparse,chan2022causal,alain2017understanding,hewitt2019designing,belrose2023eliciting}.

Aggregate faithfulness scoring runs on the 42 scored games, the T3-labeled roster minus
the 12 degenerate or static games documented in Supplement~S3, where the oracle's re-runs
are exact. The six core games of \Sectionname\ref{sec:tiers} remain the running examples
and carry the full tiered ground truth and oracle. Cell- and pixel-level scores,
correlation, deletion and insertion area-under-curve, and precision at $k$, are taken
against T1 and need no T3; only object- and concept-level scores draw on the external T3
labels, and we flag every such score as alignment-to-given, not discovery. We aggregate the
committed per-game records into one row per method on the shared
faithfulness-versus-plausibility axes of the Results leaderboard.
The headline contrast is the all-regime gap. Pooled over every output, causal and
intervention methods average a faithfulness of $0.71$ against $0.38$ for gradient and
correlational methods, a gap of $0.32$ whose $95\%$ bootstrap interval $[0.29,0.35]$
excludes zero. On the position
regime alone, where the naive gradient is zero under the hard forward map by
Prop.~\ref{prop:zero}, the two families average $0.56$ against $0.23$, a gap of $0.33$.
This gap is now statistically significant: a bootstrap over the 42 games gives a $95\%$
interval of $[0.28,0.38]$ that excludes zero. The per-method extreme places an
exact causal method, activation patching, at faithfulness $1.000$ on the position regime and
the field's default saliency at only $0.115$ there, the popular method
nonetheless carrying the higher plausibility proxy ($0.9$ versus $0.5$). We measure the
semantic gap relative to these numbers.

% =====================================================================
% BIBTEX-NEEDED: full entries for the keys cited above that are NOT yet
% in references.bib. The keys already present in references.bib that this
% section reuses are: xitari, jonas2017could, sundararajan2017axiomatic,
% selvaraju2017gradcam, chattopadhyay2018gradcampp, jaderberg2015spatial,
% jang2017gumbel, bengio2013estimating, delfosse2023ocatari.
% Please MERGE the following NEW entries (SM-only edit to references.bib):
% =====================================================================
%
% @article{maier2025vcs,
%   title   = {A bit-exact, differentiable port of the Atari 2600 Video Computer System},
%   author  = {Maier, Andreas and Bayer, Stefan and Krauss, Patrick},
%   journal = {(Paper 1; manuscript / preprint)},
%   year    = {2025},
%   note    = {jutari and jaxtari: 64/64 games bit-exact vs the xitari reference in RAM and screen}
% }
%
% @inproceedings{anand2019unsupervised,
%   title     = {Unsupervised State Representation Learning in Atari},
%   author    = {Anand, Ankesh and Racah, Evan and Ozair, Sherjil and Bengio, Yoshua and C{\^o}t{\'e}, Marc-Alexandre and Hjelm, R. Devon},
%   booktitle = {Advances in Neural Information Processing Systems (NeurIPS)},
%   year      = {2019}
% }
%
% @book{marr1982vision,
%   title     = {Vision: A Computational Investigation into the Human Representation and Processing of Visual Information},
%   author    = {Marr, David},
%   year      = {1982},
%   publisher = {W. H. Freeman and Company}
% }
%
% @inproceedings{simonyan2014deep,
%   title     = {Deep Inside Convolutional Networks: Visualising Image Classification Models and Saliency Maps},
%   author    = {Simonyan, Karen and Vedaldi, Andrea and Zisserman, Andrew},
%   booktitle = {International Conference on Learning Representations (ICLR) Workshop},
%   year      = {2014}
% }
%
% @inproceedings{zeiler2014visualizing,
%   title     = {Visualizing and Understanding Convolutional Networks},
%   author    = {Zeiler, Matthew D. and Fergus, Rob},
%   booktitle = {European Conference on Computer Vision (ECCV)},
%   year      = {2014},
%   pages     = {818--833}
% }
%
% @inproceedings{fong2017interpretable,
%   title     = {Interpretable Explanations of Black Boxes by Meaningful Perturbation},
%   author    = {Fong, Ruth C. and Vedaldi, Andrea},
%   booktitle = {IEEE International Conference on Computer Vision (ICCV)},
%   year      = {2017},
%   pages     = {3429--3437}
% }
%
% @inproceedings{petsiuk2018rise,
%   title     = {{RISE}: Randomized Input Sampling for Explanation of Black-box Models},
%   author    = {Petsiuk, Vitali and Das, Abir and Saenko, Kate},
%   booktitle = {British Machine Vision Conference (BMVC)},
%   year      = {2018}
% }
%
% @inproceedings{ribeiro2016why,
%   title     = {``Why Should I Trust You?'': Explaining the Predictions of Any Classifier},
%   author    = {Ribeiro, Marco Tulio and Singh, Sameer and Guestrin, Carlos},
%   booktitle = {Proceedings of the 22nd ACM SIGKDD International Conference on Knowledge Discovery and Data Mining},
%   year      = {2016},
%   pages     = {1135--1144}
% }
%
% @inproceedings{lundberg2017unified,
%   title     = {A Unified Approach to Interpreting Model Predictions},
%   author    = {Lundberg, Scott M. and Lee, Su-In},
%   booktitle = {Advances in Neural Information Processing Systems (NeurIPS)},
%   year      = {2017}
% }
%
% @inproceedings{abnar2020quantifying,
%   title     = {Quantifying Attention Flow in Transformers},
%   author    = {Abnar, Samira and Zuidema, Willem},
%   booktitle = {Proceedings of the 58th Annual Meeting of the Association for Computational Linguistics (ACL)},
%   year      = {2020},
%   pages     = {4190--4197}
% }
%
% @inproceedings{vig2020investigating,
%   title     = {Investigating Gender Bias in Language Models Using Causal Mediation Analysis},
%   author    = {Vig, Jesse and Gehrmann, Sebastian and Belinkov, Yonatan and Qian, Sharon and Nevo, Daniel and Singer, Yaron and Shieber, Stuart},
%   booktitle = {Advances in Neural Information Processing Systems (NeurIPS)},
%   year      = {2020}
% }
%
% @inproceedings{meng2022locating,
%   title     = {Locating and Editing Factual Associations in {GPT}},
%   author    = {Meng, Kevin and Bau, David and Andonian, Alex and Belinkov, Yonatan},
%   booktitle = {Advances in Neural Information Processing Systems (NeurIPS)},
%   year      = {2022}
% }
%
% @inproceedings{geiger2021causal,
%   title     = {Causal Abstractions of Neural Networks},
%   author    = {Geiger, Atticus and Lu, Hanson and Icard, Thomas and Potts, Christopher},
%   booktitle = {Advances in Neural Information Processing Systems (NeurIPS)},
%   year      = {2021}
% }
%
% @inproceedings{geiger2023finding,
%   title     = {Finding Alignments Between Interpretable Causal Variables and Distributed Neural Representations},
%   author    = {Geiger, Atticus and Wu, Zhengxuan and Potts, Christopher and Icard, Thomas and Goodman, Noah D.},
%   booktitle = {Proceedings of the Third Conference on Causal Learning and Reasoning (CLeaR)},
%   year      = {2024}
% }
%
% @misc{nanda2023attribution,
%   title        = {Attribution Patching: Activation Patching at Industrial Scale},
%   author       = {Nanda, Neel},
%   year         = {2023},
%   howpublished = {\url{https://www.neelnanda.io/mechanistic-interpretability/attribution-patching}}
% }
%
% @inproceedings{wang2022interpretability,
%   title     = {Interpretability in the Wild: A Circuit for Indirect Object Identification in {GPT-2} Small},
%   author    = {Wang, Kevin and Variengien, Alexandre and Conmy, Arthur and Shlegeris, Buck and Steinhardt, Jacob},
%   booktitle = {International Conference on Learning Representations (ICLR)},
%   year      = {2023}
% }
%
% @inproceedings{conmy2023towards,
%   title     = {Towards Automated Circuit Discovery for Mechanistic Interpretability},
%   author    = {Conmy, Arthur and Mavor-Parker, Augustine N. and Lynch, Aengus and Heimersheim, Stefan and Garriga-Alonso, Adri{\`a}},
%   booktitle = {Advances in Neural Information Processing Systems (NeurIPS)},
%   year      = {2023}
% }
%
% @misc{cunningham2023sparse,
%   title         = {Sparse Autoencoders Find Highly Interpretable Features in Language Models},
%   author        = {Cunningham, Hoagy and Ewart, Aidan and Riggs, Logan and Huben, Robert and Sharkey, Lee},
%   year          = {2023},
%   eprint        = {2309.08600},
%   archivePrefix = {arXiv},
%   primaryClass  = {cs.LG}
% }
%
% @misc{chan2022causal,
%   title        = {Causal Scrubbing: A Method for Rigorously Testing Interpretability Hypotheses},
%   author       = {Chan, Lawrence and Garriga-Alonso, Adri{\`a} and Goldowsky-Dill, Nicholas and Greenblatt, Ryan and Nitishinskaya, Jenny and Radhakrishnan, Ansh and Shlegeris, Buck and Thomas, Nate},
%   year         = {2022},
%   howpublished = {AI Alignment Forum}
% }
%
% @inproceedings{alain2017understanding,
%   title     = {Understanding Intermediate Layers Using Linear Classifier Probes},
%   author    = {Alain, Guillaume and Bengio, Yoshua},
%   booktitle = {International Conference on Learning Representations (ICLR) Workshop},
%   year      = {2017}
% }
%
% @inproceedings{hewitt2019designing,
%   title     = {Designing and Interpreting Probes with Control Tasks},
%   author    = {Hewitt, John and Liang, Percy},
%   booktitle = {Proceedings of the 2019 Conference on Empirical Methods in Natural Language Processing (EMNLP)},
%   year      = {2019},
%   pages     = {2733--2743}
% }
%
% @misc{belrose2023eliciting,
%   title         = {Eliciting Latent Predictions from Transformers with the Tuned Lens},
%   author        = {Belrose, Nora and Furman, Zach and Smith, Logan and Halawi, Danny and Ostrovsky, Igor and McKinney, Lev and Biderman, Stella and Steinhardt, Jacob},
%   year          = {2023},
%   eprint        = {2303.08112},
%   archivePrefix = {arXiv},
%   primaryClass  = {cs.LG}
% }
