% 04_results_A.tex — Results A: the neuroscience battery, quantified (P2-E8-5).
% Maier voice per ../STYLE.md. Every number traces to a committed §R record
% (tools/xai_study/phaseA_kording/out/*.json) via the E6-1 leaderboard and Fig. 3.
% Honesty contract (plan.md): A is the low-faithfulness calibration baseline, not the
% headline; we report what the methods recover against the KNOWN mechanism, and we do
% NOT claim any of them recovered meaning or algorithm.

\subsection{The neuroscience battery}\label{sec:results-a}

We begin with a known concern in the field. Jonas and Kording asked whether a
neuroscientist with the standard toolkit could understand a microprocessor. They
found that the toolkit yields rich structure but misses the
mechanism~\cite{jonas2017could}. They could show the failure but not measure it.
On the 6502 they had no causal ground truth to score each method against. We
supply that missing number. We run the same battery on the bit-exact, fully
known VCS, where exact intervention recovers the true causal role of every
register and RAM cell (\Sectionname\ref{sec:methods}). This makes Kording's
qualitative lesson quantitative. Across the eight analyses A1--A8,
faithfulness against the oracle ranges from $0.05$ to $1.0$ and averages
$0.45$~\cite{leaderboardP2}. The
oracle-as-method positive
control sits at $1.0$ by construction (Table~\ref{tab:phaseA}). The structure
these methods report is real. But structure is not the same as cause. The score
measures the difference.

The dependency structure is largely recoverable. A1 reconstructs the dependency
graph over state variables, a connectome of read/write edges, and scores it
against the true data-flow graph at a faithfulness of
$0.34$~\cite{leaderboardP2}: most of its edges are spurious. The single-unit
lesion A2 fares best. It clamps each RAM cell in turn and ranks it by the
behavioral damage it does. The resulting importance map has a Spearman
correlation of $0.99$ with the unit's true causal role (mean $0.992$ over the
six core games)~\cite{leaderboardP2}. Lesioning is causal, which is why A2 is
the one classical method that scores near the top.

Yet a near-perfect rank order still misses interactions. A single-unit lesion
changes one cell at a time, so it is blind to causes that act only together. On
Space Invaders the per-unit ranking misses $40\%$ of the interactions the oracle
records. One super-additive pair swings behavior by up to $1044$ screen-pixels
beyond the sum of its parts~\cite{leaderboardP2}. The ranking of which units
matter is correct, but the model of how they combine is missing. So even the
strongest classical method recovers a ranking, not the computation. An effect
map does not show the algorithm behind it.

Tuning curves report structure that does not match the true cause. A3 fits
per-unit tuning to a game variable, the workhorse of systems neuroscience. On the
VCS it actively misleads. Its \emph{spurious-tuning rate} is the fraction of
strongly-tuned units whose tuning does not match their true causal role. That
rate is $0.50$ on Pong, and its faithfulness averages $0.09$, the lowest in the
battery~\cite{leaderboardP2}. Tuning is correlational. On a machine where many
cells co-vary with the beam clock, apparent tuning is cheap. This is the same
present-versus-used distinction we discuss later for linear probing.

Correlation structure and field potentials are real, but they are mostly side
effects, not causes. A4, the spike-word analysis of pairwise and global
correlation, reproduces the familiar weak-pairwise/strong-global signature. Yet
against the true coupling its faithfulness is only $0.22$, with sufficiency near
zero ($S = 0.026$)~\cite{leaderboardP2}. A5 is the local-field-potential
analogue that pools regional activity into power spectra. It scores $0.30$, and
much of that variance is the known frame and scanline
clocks~\cite{leaderboardP2}. Both methods detect the machine's activity but not
its computation.

Granger causality treats ``happens earlier'' as ``is the cause.'' A6 infers
directed CPU/TIA/RIOT edges and scores them against the true data-flow. On Pong
its faithfulness is $0.0$, with false-edge and missed-edge rates both at
$1.0$~\cite{leaderboardP2}. The VCS is clock-locked, so nearly everything is
predictable from everything else one cycle earlier. Equating precedence with
causation therefore infers spurious edges everywhere. Neither a longer lag nor a
stricter threshold fixes this. Precedence and causation differ here, and only
the intervention oracle separates them.

Dimensionality reduction scores in the middle, with NMF and PCA tied. A7
decomposes the state tensor into latent components and matches them to the known
clock, read/write, and vsync signals. Non-negative matrix factorization and PCA
both match about 60 percent of the known components on average (at best 80 percent
on a single game), so neither factorization is the more faithful at this
analogue~\cite{leaderboardP2}. NMF leads only on the secondary recon-error/FSM
composite, where its non-negativity better fits the additive register basis.
This makes A7 the second-highest score, behind only the causal lesion. It still
leaves 40 percent of the components unmatched.

A complete record is still not an explanation. A8 records the full RAM and
register state, faithful and sufficient by definition ($F = 1.0$, $S = 1.0$).
Yet its minimality is only $0.08$~\cite{leaderboardP2}. At the algorithmic level,
a recording of every cell explains nothing. Faithfulness and sufficiency without
minimality give a log file, not an account of the computation. This is the first
sign that the three axes are independent. The semantic-gap result makes that
point central later.

Taken together, the battery confirms the expected result and puts a number on
it. The advance over Jonas and Kording is that we grade the \emph{architectural}
register-transfer level with a \emph{quantitative} $F$ against an exact oracle,
rather than giving a qualitative verdict on the transistor. The contrast is
clear, because the causal operators we apply to the same machine in
\Sectionname\ref{sec:results-b} and~\Sectionname\ref{sec:results-c} reach the ceiling this battery
cannot. So the limitation lies in the analysis methods, not in the machine. We
mean this strictly as calibration. Even the battery's best members recover a
ranking, a graph, or a record, never the algorithm the program runs. This
remaining gap is what the paper measures.

\begin{table}[t]
\footnotesize
\setlength{\tabcolsep}{4pt}
\caption{\textbf{The neuroscience battery scored against the intervention oracle.}
Faithfulness $F$ of each classical analysis A1--A8 against the exact intervention oracle
(\Sectionname\ref{sec:oracle}), with its $95\%$ bootstrap confidence interval over the six
core games and the documented plausibility proxy (\Sectionname\ref{sec:methods}). The battery
averages $0.45$; the oracle-as-method positive control is $1.0$ by construction. Only the
causal single-unit lesion (A2) reaches the ceiling on a non-trivial metric. The whole-state
recording (A8) is at $1.0$ only because recording everything is faithful, yet its minimality
is $0.07$ --- a log file, not an account. Every value reads from a committed per-game record
via the leaderboard~\cite{leaderboardP2}; none is re-run or estimated.}
\label{tab:phaseA}
\begin{tabularx}{\textwidth}{@{}l c c >{\raggedright\arraybackslash}X@{}}
\toprule
Analysis & $F$ ($95\%$ CI) & Plaus & What $F$ scores \\
\midrule
A1 connectomics             & $0.34$ $[0.09,0.67]$ & $0.55$ & data-flow-graph $F_1$ vs.\ oracle \\
A2 single-unit lesions      & $0.99$ $[0.99,1.00]$ & $0.55$ & lesion-importance $\rho$ vs.\ true role \\
A3 tuning curves            & $0.09$ $[0.00,0.25]$ & $0.80$ & $1-$spurious-tuning rate \\
A4 spike-word correlations  & $0.22$ $[0.07,0.41]$ & $0.80$ & coupling vs.\ true coupling \\
A5 local field potentials   & $0.30$ $[0.22,0.39]$ & $0.80$ & $1-$clock-explained variance \\
A6 Granger causality        & $0.05$ $[0.00,0.11]$ & $0.80$ & $1-$false-edge rate vs.\ data-flow \\
A7 dimensionality reduction & $0.60$ $[0.60,0.60]$ & $0.75$ & NMF/PCA matched-component fraction \\
A8 whole-state recording    & $1.00$ $[1.00,1.00]$ & $0.30$ & complete record (minimality $M=0.08$) \\
\midrule
Oracle (positive control)   & $1.00$ $[1.00,1.00]$ & $1.00$ & exact intervention ground truth \\
\botrule
\end{tabularx}
\end{table}

% BIBTEX-NEEDED: full entries for keys cited above that are NOT already in
% references.bib. (jonas2017could and sundararajan2017axiomatic already exist.)
% Please merge the following into references.bib at SM Review.
%
% @misc{leaderboardP2,
%   author       = {{Paper~2 authors}},
%   title        = {Cross-tradition faithfulness--plausibility leaderboard for the
%                   differentiable {Atari} {VCS} interpretability audit (Phases A--C)},
%   year         = {2026},
%   howpublished = {Committed analysis record,
%                   \texttt{tools/xai\_study/compare/out/leaderboard.json};
%                   aggregates the per-game §R records under
%                   \texttt{tools/xai\_study/phaseA\_kording/out/}},
%   note         = {Internal data artifact; 257 per-game records aggregated into 31
%                   method rows including the oracle positive control. Generated by
%                   \texttt{compare/leaderboard.py}. Accessed 2026-06-23.}
% }
