% 06_results_C.tex — Results C: mechanistic interpretability (P2-E8-7).
% Phase-C: the mech-interp toolkit scored against the exact patch / true data-flow /
% scrubbing oracle, on a known real circuit. Maier voice (../STYLE.md). Numbers are
% committed §R values (leaderboard.md + phaseC_mechanistic/out/*_core_summary.json).

\subsection{Mechanistic interpretability on a known circuit}\label{sec:results-c}

Mechanistic interpretability makes the strongest claim in the field. It tries to explain how
a system computes, not just where it attends. Its methods read internal activations as a circuit and
try to recover that circuit. On a neural network the recovered circuit can only be checked
against itself, because the true wiring is unknown. The VCS removes that limitation. Its
data-flow is the program's data-flow, known exactly from the disassembly. Every
intermediate value is a RAM cell or a register we can read, clamp, and re-run bit-exactly.
We therefore run ten mechanistic-interpretability methods on a circuit whose answer we
already know. We score each against the same \Sectionname\ref{sec:oracle}~oracle: the exact
intervention table, the true read/write graph, and the F\,$\wedge$\,S\,$\wedge$\,M triad of
\Sectionname\ref{sec:methods}. This calibrates the
toolkit against a \emph{known} circuit inside a real, human-written artifact rather than a
synthetic or compiled one. Table~\ref{tab:phaseC} places the mechanistic methods against the
oracle. The result splits cleanly. The causal methods are exact, but their
exactness recovers the wiring of the circuit without recovering what the variables mean.

\begin{table}[t]
\footnotesize
\setlength{\tabcolsep}{4pt}
\caption{\textbf{Mechanistic interpretability scored against the intervention oracle
(\Sectionname\ref{sec:oracle}).} Ten methods aggregated over the 42 scored games (oracle-as-method positive
control $=1.0$): faithfulness $F$ with its $95\%$ bootstrap CI, the plausibility proxy
(\Sectionname\ref{sec:methods}), and the kind of recovery the oracle certifies. The causal core is
exact by construction; the approximate and search-based methods lose recall; and the
named-but-unused and present-but-unused methods sit in the danger zone (high plausibility, low
faithfulness). Values read from the committed leaderboard~\cite{leaderboardP2}.}
\label{tab:phaseC}
\begin{tabularx}{\textwidth}{@{}X l c >{\raggedright\arraybackslash}X@{}}
\toprule
Method & $F$ ($95\%$ CI) & Plaus & Recovery \\
\midrule
Activation patching~\cite{vig2020investigating}          & $1.00$ $[1.00,1.00]$ & $0.50$ & exact effect table \\
Interchange interventions (DAS)~\cite{geiger2023finding} & $1.00$ $[1.00,1.00]$ & $0.50$ & verifies a supplied alignment \\
Logit / tuned lens~\cite{belrose2023eliciting}           & $0.82$ $[0.77,0.87]$ & $0.50$ & exact intermediate value \\
Causal scrubbing~\cite{chan2022causal}                   & $0.98$ $[0.94,1.00]$ & $0.50$ & exact data-flow preserved \\
ACDC~\cite{conmy2023towards}                              & $0.63$ $[0.52,0.74]$ & $0.50$ & search (graph-correct, $S{=}0.30$) \\
Path patching~\cite{wang2022interpretability}            & $0.58$ $[0.45,0.71]$ & $0.50$ & search (single-shot edges) \\
Attribution patching~\cite{nanda2023attribution}         & $0.46$ $[0.34,0.57]$ & $0.90$ & approximate (linear surrogate) \\
NMF/PCA dictionaries                                     & $0.10$ $[0.05,0.16]$ & $0.75$ & named, partly unused \\
Sparse autoencoder~\cite{cunningham2023sparse}           & $0.15$ $[0.10,0.22]$ & $0.75$ & named yet unused ($F{=}0.04$ on Pong) \\
Linear probing ($+$ control)~\cite{hewitt2019designing}  & $0.24$ $[0.15,0.35]$ & $0.85$ & present, not used \\
\midrule
Oracle (positive control)                               & $1.00$ $[1.00,1.00]$ & $1.00$ & exact intervention ground truth \\
\botrule
\end{tabularx}
\end{table}

\subsubsection*{Exact recovery of the wiring}

The main result is simple. Activation patching recovers the true causal-effect table
exactly. Across all six core games the recovered per-site effect equals the exact
intervention oracle to the last bit (maximum
absolute deviation $0.0$), with site precision and recall both $1.0$. On this machine,
clamping a RAM cell to a donor value and re-running the real ROM is literally the oracle's
own operation. So it cannot disagree with the oracle. Causal scrubbing inherits the same
strength. The true data-flow hypothesis preserves behaviour under resample-ablation of
everything outside it on all six core games (preserved performance $1.0$), and averages
$0.98$ over the scored games. A deliberately wrong
hypothesis does not, and the two are discriminated on every core game. Interchange
interventions with distributed alignment search (DAS) verify the supplied alignment at
interchange accuracy $1.0$ against a mis-aligned control at $0.0$. The logit lens reads
the true intermediate at readout fidelity $1.0$ on every game, yet only $0.82$ of its
readout weight lands on causally-relevant sites --- its faithfulness. Two of the
causal methods sit at faithfulness $1.0$, level with the oracle positive control
(activation patching and DAS), and
causal scrubbing at $0.98$ just below it
(Table~\ref{tab:phaseC}). Where a method's operation \emph{is} an intervention, faithfulness
is automatic on a fully observable machine. This is the first part of the problem, and it is
solved here.

\subsubsection*{Approximate and search-based recovery}

The causal account weakens as soon as a method replaces exact re-runs with an approximation
or a search. Attribution patching, the gradient-linear surrogate for activation patching,
correlates with the exact patch at only $0.50$ and matches it inside the linear regime in
just $0.17$ of cases. Yet its recovered edges stay precise (edge precision $0.98$, recall
$0.70$) because the linearisation is wrong about the size of an edge, not about whether the
edge exists. Path patching,
run in the IOI style as a single directed do-operation, recovers the directed edge set of
the true routine at mean $F1=0.25$. On Q*bert it matches the empty direct circuit perfectly
($F1=1.0$). On Pong it recovers none of the eleven true edges ($F1=0.0$), because each
receiver's value is re-derived within the horizon, so no one-step path carries a surviving
signal. ACDC discovers the circuit automatically by resample-ablation edge pruning over a
threshold sweep. It reaches mean $F1=0.47$ (ROC-AUC $0.70$). A positive control re-runs the
identical edge-restricted procedure under the exhaustive do-set and scores $F1=1.0$ on every
game, so the deficit is a real measurement of single-shot recovery, not a broken scorer.
These are the methods the field reaches for when the exact intervention is too expensive.
On a machine where the exact intervention is cheap, these methods miss real structure. They
are precise about what they find but silent about dependencies re-derived through other
cells. The limit is recall, not precision. The gap to the positive control is the quantified
cost of approximation against a circuit that is known.

\subsubsection*{Three separations from the oracle}

This subsection presents the central result. ACDC on Q*bert recovers the data-flow circuit
at perfect $F1$
against the true routine, and the recovered graph is minimal at the algorithmic level
(F$=1.0$, M$=1.0$). Yet that same graph, scrubbed and asked to regenerate the behaviour
under held-out interventions, preserves only S$=0.10$ of it (mean scrub-preserved
performance $0.30$ across games). A circuit can have the correct graph and still not
reproduce the computed function. The missing object is not a denser or cleaner graph of the
same type. Activation patching has already recovered the effect table exactly, with
no tuning budget left to spend. A table of per-site effects is still not the bounce algorithm. One
method is exact and the other recovers a perfect graph. Even so, neither reproduces the
behaviour. A ``methods underperform'' explanation cannot account for this separation, because
no method underperformed here.

The dictionary methods show the same gap from the opposite direction. A sparse autoencoder
trained on
the recorded RAM trajectory reconstructs the state almost perfectly (fraction of variance
explained $>0.99$ on Pong) and matches every verified game-variable cell it was asked to
(matched fraction $1.0$). By the metric the field uses to claim a feature ``is'' a concept,
the SAE succeeds. By the oracle, it does not: ablating the matched Pong feature
and re-running the real ROM gives causal faithfulness F$=0.041$, with S$=-0.46$. When we
remove the matched feature, the computation it is supposed to name does not change. Across
the scored
games the SAE's mean leaderboard faithfulness is $0.17$ against a method-tradition
plausibility proxy of $0.75$, squarely in the \Sectionname\ref{sec:methods} danger zone. The
NMF/PCA dictionary contrast repeats the pattern at lower resolution (matched-component
fractions up to $0.75$ beside near-zero causal use). This separation is the reverse of the
first. The circuit was causal but behaviourally incomplete; the feature is perfectly
named but never used. The two failures point in opposite directions.

Linear probing shows a third version of the same gap, the well-known probing trap. With
control tasks (Hewitt \&
Liang's selectivity), probes decode game-concept cells well above chance (mean probe accuracy
$0.95$, mean selectivity $0.11$), confirming that the information is present in the state.
Yet the program does not necessarily use that information. Cross-referenced against the exact
intervention oracle, the oracle
flags $14$ concept cells across the core games that the program never causally uses within the
window. Of those, $2$ are also decodable above chance, so a probe would have nominated them as
encoded concepts. Breakout's score cell~84 sits in the larger set: it is present in the state
and linearly associated with the displayed score, yet provably unused under intervention within
the window.
The oracle is strictly stronger than the probe. The probe asks whether a value can be read
out. The oracle asks whether changing that value changes the output. Probing earns a high plausibility proxy ($0.85$)
on a faithfulness of $0.24$, squarely in the danger zone (Table~\ref{tab:phaseC}).
We state the caveat directly: these labels are imported externally (AtariARI,
OCAtari) and only \emph{verified} causally, never \emph{discovered}. So DAS's alignment
accuracy of $1.0$ is guaranteed in advance, because it only confirms a variable we supplied.
The probe tells us a concept is present. Only the intervention tells us the program uses it.

\subsubsection*{Exactness is necessary but not sufficient}

Mechanistic interpretability is the clearest test of the paper's claim. Run against a known real circuit, the
mechanistic toolkit splits into two groups. The first is a causal core that is faithful by
construction: activation patching and DAS at $1.0$, and the logit lens at $0.82$, with causal scrubbing
at $0.98$. Together with the intervention methods this causal and intervention family averages
$0.71$ faithfulness, the highest of the three traditions. The second is a recovery
layer that is either precise but partial (attribution and path patching, ACDC) or present but
unused (SAEs, dictionaries, probing). The oracle, not our judgement, decides the three
separations. They appear exactly where the field's success metrics report success:
graph-correct yet behaviour-incomplete (ACDC on Q*bert F$=1.0$/M$=1.0$, S$=0.10$), named yet
causally unused (SAE on Pong matched $1.0$, F$=0.041$), present yet unused (cell~84). The
faithfulness advantage that the causal and intervention family holds over the gradient
methods is robust across all output regimes. The all-regime gap is $0.32$ (95\% CI
$[0.29,0.35]$, excluding zero). Restricted to the position-output regime alone the gap is
$0.33$, in the same direction, and its $95\%$ bootstrap interval over the 42 games
$[0.28,0.38]$ now excludes zero, so on position the causal advantage is statistically
significant, no longer only directional. The
separations above do not rest on either family mean, because each is a within-method
contradiction on a single game. On a machine that gives faithfulness for free, faithfulness is necessary for
understanding but not sufficient. The residual is the semantic gap: the wiring is recovered,
but the meaning is not. \Sectionname\ref{sec:results_compare} places this gap on the shared axes, and
the Discussion treats it as the field's open problem.

% =====================================================================
% BIBTEX-NEEDED: new keys cited above, for the SM to merge into references.bib.
% (Do NOT edit references.bib here — SM-only.)
% =====================================================================
% @inproceedings{vig2020causalmediation,
%   author    = {Vig, Jesse and Gehrmann, Sebastian and Belinkov, Yonatan and Qian, Sharon and Nevo, Daniel and Singer, Yaron and Shieber, Stuart},
%   title     = {Investigating Gender Bias in Language Models Using Causal Mediation Analysis},
%   booktitle = {Advances in Neural Information Processing Systems (NeurIPS)},
%   volume    = {33},
%   pages     = {12388--12401},
%   year      = {2020}
% }
% @inproceedings{meng2022rome,
%   author    = {Meng, Kevin and Bau, David and Andonian, Alex and Belinkov, Yonatan},
%   title     = {Locating and Editing Factual Associations in {GPT}},
%   booktitle = {Advances in Neural Information Processing Systems (NeurIPS)},
%   volume    = {35},
%   pages     = {17359--17372},
%   year      = {2022}
% }
% @inproceedings{geiger2021abstraction,
%   author    = {Geiger, Atticus and Lu, Hanson and Icard, Thomas and Potts, Christopher},
%   title     = {Causal Abstractions of Neural Networks},
%   booktitle = {Advances in Neural Information Processing Systems (NeurIPS)},
%   volume    = {34},
%   pages     = {9574--9586},
%   year      = {2021}
% }
% @inproceedings{geiger2023das,
%   author    = {Geiger, Atticus and Wu, Zhengxuan and Potts, Christopher and Icard, Thomas and Goodman, Noah D.},
%   title     = {Finding Alignments Between Interpretable Causal Variables and Distributed Neural Representations},
%   booktitle = {Proceedings of the Third Conference on Causal Learning and Reasoning (CLeaR)},
%   year      = {2024},
%   note      = {arXiv:2303.02536}
% }
% @inproceedings{conmy2023acdc,
%   author    = {Conmy, Arthur and Mavor-Parker, Augustine N. and Lynch, Aengus and Heimersheim, Stefan and Garriga-Alonso, Adri{\`a}},
%   title     = {Towards Automated Circuit Discovery for Mechanistic Interpretability},
%   booktitle = {Advances in Neural Information Processing Systems (NeurIPS)},
%   volume    = {36},
%   year      = {2023},
%   note      = {arXiv:2304.14997}
% }
% @inproceedings{wang2022ioi,
%   author    = {Wang, Kevin and Variengien, Alexandre and Conmy, Arthur and Shlegeris, Buck and Steinhardt, Jacob},
%   title     = {Interpretability in the Wild: A Circuit for Indirect Object Identification in {GPT-2} Small},
%   booktitle = {International Conference on Learning Representations (ICLR)},
%   year      = {2023},
%   note      = {arXiv:2211.00593}
% }
% @article{goldowskydill2023pathpatching,
%   author    = {Goldowsky-Dill, Nicholas and MacLeod, Chris and Sato, Lucas and Arora, Aryaman},
%   title     = {Localizing Model Behavior with Path Patching},
%   journal   = {arXiv preprint arXiv:2304.05969},
%   year      = {2023}
% }
% @misc{nanda2023attributionpatching,
%   author    = {Nanda, Neel},
%   title     = {Attribution Patching: Activation Patching at Industrial Scale},
%   howpublished = {\url{https://www.neelnanda.io/mechanistic-interpretability/attribution-patching}},
%   year      = {2023}
% }
% @article{syed2023attributionpatching,
%   author    = {Syed, Aaquib and Rager, Can and Conmy, Arthur},
%   title     = {Attribution Patching Outperforms Automated Circuit Discovery},
%   journal   = {arXiv preprint arXiv:2310.10348},
%   year      = {2023}
% }
% @article{cunningham2023sparse,
%   author    = {Cunningham, Hoagy and Ewart, Aidan and Riggs, Logan and Huben, Robert and Sharkey, Lee},
%   title     = {Sparse Autoencoders Find Highly Interpretable Features in Language Models},
%   journal   = {arXiv preprint arXiv:2309.08600},
%   year      = {2023}
% }
% @article{bricken2023monosemanticity,
%   author    = {Bricken, Trenton and Templeton, Adly and Batson, Joshua and Chen, Brian and Jermyn, Adam and others},
%   title     = {Towards Monosemanticity: Decomposing Language Models with Dictionary Learning},
%   journal   = {Transformer Circuits Thread},
%   year      = {2023},
%   note      = {\url{https://transformer-circuits.pub/2023/monosemantic-features}}
% }
% @article{templeton2024scaling,
%   author    = {Templeton, Adly and Conerly, Tom and Marcus, Jonathan and Lindsey, Jack and Bricken, Trenton and others},
%   title     = {Scaling Monosemanticity: Extracting Interpretable Features from {Claude 3 Sonnet}},
%   journal   = {Transformer Circuits Thread},
%   year      = {2024},
%   note      = {\url{https://transformer-circuits.pub/2024/scaling-monosemanticity}}
% }
% @misc{chan2022causalscrubbing,
%   author    = {Chan, Lawrence and Garriga-Alonso, Adri{\`a} and Goldowsky-Dill, Nicholas and Greenblatt, Ryan and Nitishinskaya, Jenny and Radhakrishnan, Ansh and Shlegeris, Buck and Thomas, Nate},
%   title     = {Causal Scrubbing: A Method for Rigorously Testing Interpretability Hypotheses},
%   howpublished = {AI Alignment Forum},
%   year      = {2022}
% }
% @inproceedings{alain2017probing,
%   author    = {Alain, Guillaume and Bengio, Yoshua},
%   title     = {Understanding Intermediate Layers Using Linear Classifier Probes},
%   booktitle = {International Conference on Learning Representations (ICLR) Workshop},
%   year      = {2017},
%   note      = {arXiv:1610.01644}
% }
% @inproceedings{hewitt2019control,
%   author    = {Hewitt, John and Liang, Percy},
%   title     = {Designing and Interpreting Probes with Control Tasks},
%   booktitle = {Proceedings of the 2019 Conference on Empirical Methods in Natural Language Processing (EMNLP-IJCNLP)},
%   pages     = {2733--2743},
%   year      = {2019}
% }
% @misc{nostalgebraist2020logitlens,
%   author    = {{nostalgebraist}},
%   title     = {Interpreting {GPT}: The Logit Lens},
%   howpublished = {LessWrong},
%   year      = {2020},
%   note      = {\url{https://www.lesswrong.com/posts/AcKRB8wDpdaN6v6ru/interpreting-gpt-the-logit-lens}}
% }
% @article{belrose2023tunedlens,
%   author    = {Belrose, Nora and Furman, Zach and Smith, Logan and Halawi, Danny and Ostrovsky, Igor and McKinney, Lev and Biderman, Stella and Steinhardt, Jacob},
%   title     = {Eliciting Latent Predictions from Transformers with the Tuned Lens},
%   journal   = {arXiv preprint arXiv:2303.08112},
%   year      = {2023}
% }
% @inproceedings{anand2019atariari,
%   author    = {Anand, Ankesh and Racah, Evan and Ozair, Sherjil and Bengio, Yoshua and C{\^o}t{\'e}, Marc-Alexandre and Hjelm, R. Devon},
%   title     = {Unsupervised State Representation Learning in {Atari}},
%   booktitle = {Advances in Neural Information Processing Systems (NeurIPS)},
%   volume    = {32},
%   year      = {2019}
% }
