% 09_endmatter.tex — End matter (data/code availability, author contributions,
% competing interests, ethics/broader impact, acknowledgements). Drafted by item
% P2-E8-9 (P2 Sprint 7); revised by P2-R-S09-endmatter (P2 Revision Sprint R5).
% Voice: ../STYLE.md (Andreas Maier). Requirements: document_check.md §D.
% Availability statements kept consistent with the repro bundle (REPRODUCIBILITY.md
% §0). Per revision_plan.md PO decision #3 (verified against NMI policy 2026-06):
% the XAI benchmark/scoring suite is "available for review" NOW (single-blind, public
% GitHub link), with a Zenodo DOI snapshot for the published Code Availability statement
% and an optional Code Ocean capsule for code peer review — NOT "released upon
% acceptance". The companion emulator substrate is already public (arXiv:2606.22447;
% github.com/akmaier/UnderstandingVCS). ROMs are not redistributed (AutoROM + SHA-256
% hash table in the SI). The companion-paper roadmap is a brief forward pointer only
% and does not lean on unpublished Papers 4/5 (improvement P3#37).
%
% P2-E9-1 (P2 Sprint 8) document_check pass: verified that every NMI-required
% back-matter statement is present here as flowing \bmhead prose with NO \paragraph
% run-in labels (STYLE.md §1/§3) — Data availability, Code availability, Author
% contributions, Competing interests (none), Ethics/broader impact. The Limitations
% paragraph is NOT duplicated here; it lives in 08_discussion.tex (the "several
% limits bound what we may conclude" paragraph, as ordinary prose per STYLE.md §5).
% Per-requirement mapping recorded in ../document_check.md.

\backmatter

\bmhead{Data availability}
The interpretability scores reported here come from the committed analysis
records under \texttt{tools/xai\_study/}. These records hold the per-method, per-game §R
results and the aggregated cross-tradition leaderboard~\cite{leaderboardP2,faithfuldemoP2}.
Every quantitative claim in the Article reads from one of these committed artifacts. No
number is re-run or estimated. The game cartridges used as the subject are
third-party Atari~2600 ROMs that we do \emph{not} redistribute. We follow the
practice of the underlying emulator port~\cite{maier2025vcs}. We acquire the ROMs through
the \texttt{AutoROM} retrieval procedure and report a SHA-256 hash for each ROM
in the Supplementary Information. This lets others reproduce and verify the exact binaries
independently. The recorded state trajectories used to
train the sparse autoencoders and to run the patching and probing experiments are
regenerated deterministically from those ROMs and the published action streams.

\bmhead{Code availability}
The interpretability benchmark introduced here is available for review at
\url{https://github.com/akmaier/UnderstandingVCS}. It comprises the scoring suite, the
F\,$\wedge$\,S\,$\wedge$\,M triad metrics, the intervention oracle, and the per-phase
method implementations (\texttt{tools/xai\_study/}). The repository also holds the
reproducibility appendix that maps every reported number to its committed record and run
command. A versioned snapshot will be archived on Zenodo with a minted DOI. An optional
Code Ocean capsule reproduces the leaderboard and the figures from the committed result
records alone. The benchmark rests on a differentiable, bit-exact emulator. It has two
interchangeable versions (\texttt{jutari} in Julia, \texttt{jaxtari} in JAX/Python), and
both are already public. The emulator is described in the companion port
paper~\cite{maier2025vcs} (arXiv:2606.22447), and its source is openly available at the
same repository. The companion paper supplies the bit-exactness validation and the
gradient caveat that the present work imports. What is new here is the interpretability
benchmark, the oracle scoring, and the semantic-gap diagnosis. We release the code
under a permissive open-source licence so that a third party can score a new method
against the same ground-truth oracle.

\bmhead{Author contributions}
A.M. conceived the study, designed the ground-truth audit and the
F\,$\wedge$\,S\,$\wedge$\,M triad, implemented the benchmark and oracle scoring, and
wrote the manuscript. S.B. contributed to the method design and the analysis and
revised the manuscript. P.K. contributed to the differentiable substrate and oracle
and to the gradient-faithfulness analysis, and helped develop the differentiability
framing. All authors discussed the results, interpreted the semantic-gap finding, and
approved the final manuscript.

\bmhead{Competing interests}
The authors declare no competing interests.

\bmhead{Ethics and broader impact}
This study uses no human-subjects data; the plausibility axis is a documented
plausibility proxy by method tradition rather than a measurement, and we report it as
such. The subject is a fully deterministic, hand-coded program with no learned policy, so no
agent is trained. Interpretability methods are dual-use: a tool that audits a
system can also be used to probe it. Our contribution is a calibration that makes
explanation claims more trustworthy, and we release it openly to that end.
The game cartridges are not redistributed.

\bmhead{Acknowledgements}
The authors thank the maintainers of the AtariARI~\cite{anand2019unsupervised} and
OCAtari~\cite{delfosse2023ocatari} projects, whose game-concept labels we imported as
external candidates and only causally verified, and the maintainers of the
\texttt{xitari} reference emulator~\cite{xitari} against which the substrate is
validated bit-for-bit. The pilots were run locally; the full sweeps used the
institutional high-performance compute cluster of Friedrich-Alexander-Universit\"at
Erlangen-N\"urnberg. This work received no specific external funding.

% =====================================================================
% BIBTEX-NEEDED: this section introduces NO new keys. Every key cited above is
% already declared in a sibling section's BIBTEX-NEEDED block (SM to merge):
%   leaderboardP2          -> 04_results_A.tex
%   faithfuldemoP2         -> 07_results_compare.tex
%   maier2025vcs, xitari   -> 03_methods.tex
%   anand2019unsupervised  -> 03_methods.tex
%   delfosse2023ocatari    -> 03_methods.tex
% =====================================================================
