\section{Coverage and applicability}\label{supp:coverage}

The benchmark scores methods on a six-game core drawn from a larger bit-exact substrate. The
coverage table makes the scope of that choice explicit. Paper~1 validates the
differentiable VCS ports bit-for-bit against the \texttt{xitari} reference on all 64 Arcade
Learning Environment games, within a 30-frame RAM and 60-frame screen window. The
interpretability study reported here runs on the six games listed in
Table~\ref{supp:tab:coverage}. Every method is scored on a shared gameplay state, produced by
one seeded random-action stream (seed~$0$, $90$-frame prefix, $15$-frame analysis horizon), not
on the boot screen. This keeps the analysis inside the Paper~1 bit-exact window while giving the
oracle a rich, non-degenerate cause set to score against. We pick these six games because all
three tiers of ground truth (\Sectionname\ref{sec:tiers}) are available for them and the oracle is
exact. The remaining 58 games are
bit-exact substrate but are not scored here. Some fall outside the validated horizon for the
analyses we run. For others, the tiered labels they would need are not yet produced. They are
candidates for the larger sweep. \Sectionname\ref{supp:t3coverage}
(Table~\ref{supp:tab:t3coverage}) records how far the external T3 labels reach across the
64-game roster, defining a 54-game labeled breadth set from which that sweep will draw; the
six games here are the core subset of it that carries the full tiered ground truth. Games are
not excluded for any methodological reason. The
horizon column records the validated window. The tier column records which of the three
ground-truth tiers (T1 exact intervention, T2 data-flow, T3 externally-supplied semantic
labels) is available.

\begin{table}[t]
\centering
\caption{\textbf{Core-game coverage.} The six core games carry exact tiered ground truth and
the full method battery. Faithfulness is scored within the Paper~1 bit-exact window. The
last two columns give the methods run and the per-game output records contributed. The wider
64-game substrate is bit-exact (Paper~1) but is reserved for the larger sweep. T3 labels are
external (\Sectionname\ref{sec:tiers}); we record this so the benchmark does not appear to certify them.}
\label{supp:tab:coverage}
\footnotesize
\setlength{\tabcolsep}{4pt}
\begin{tabular}{@{}p{2.4cm}p{1.9cm}p{1.5cm}p{1.7cm}p{1.5cm}p{1.4cm}@{}}
\toprule
Game & Paper-1 status & Horizon & Tiers & Methods run & Records \\
\midrule
breakout & bit-exact & 30/60 fr & T1,T2,T3 & 30 + oracle & per-game \\
ms\_pacman & bit-exact & 30/60 fr & T1,T2,T3 & 30 + oracle & per-game \\
pong & bit-exact & 30/60 fr & T1,T2,T3 & 30 + oracle & per-game \\
qbert & bit-exact & 30/60 fr & T1,T2,T3 & 30 + oracle & per-game \\
seaquest & bit-exact & 30/60 fr & T1,T2,T3 & 30 + oracle & per-game \\
space\_invaders & bit-exact & 30/60 fr & T1,T2,T3 & 30 + oracle & per-game \\
\midrule
58 further games & bit-exact (P1) & 30/60 fr & T1,T2 partial & reserved & --- \\
\bottomrule
\end{tabular}
\end{table}

A gameplay state can still be degenerate for a particular method, and the benchmark records
this rather than scoring it as a failure. The oracle cause-density gate is the first guard. It
counts how many candidate causes move the shared output above a floor, and rejects a state that
does not clear a minimum count, so a near-flat oracle column never reaches a method. On
\texttt{seaquest} the gate raises the count of live candidate causes at the analysis frame from
$2$ to $19$, which is what makes the game usable. Three residual degenerate cells survive the
gate and are marked, not scored as zero. On \texttt{ms\_pacman} the pairwise neuroscience
analysis (A4) finds no coupled candidate pair at that frame, so its cell is uninformative. The
on-distribution counterfactual returns a zero delta on \texttt{space\_invaders},
\texttt{ms\_pacman}, and \texttt{qbert}, where its perturbation does not move the shared output;
those cells are recorded as not-valid. Expected Gradients scores low on the content regime of
every game but \texttt{pong}, because the causal byte it should attribute to is static there.
None of these cells is read as a wrong answer.

Whether a method applies at all depends on what the method needs from its target. The
applicability matrix records that for every method family. The Phase~A neuroscience analyses
read the raw machine state and need neither a learned policy nor gradients. The Phase~B
attribution methods split into two groups. The gradient members need a differentiable forward
pass. The intervention members need only the ability to set state. The Phase~C mechanistic
methods need access to internal activations. The causal members among them also need the
ability to intervene. Table~\ref{supp:tab:applic} records, per family, whether the method
operates on the raw VCS, whether it requires a learned policy, gradients, internal layers or
attention, or external labels, whether it uses oracle-like interventions, and whether it
receives a supplied hypothesis. The table shows the same pattern the results confirm. The
families that score high on faithfulness are the ones that can intervene. The families that
collapse on the position regime are the gradient and correlational ones.

\begin{table}[t]
\centering
\caption{\textbf{Per-family applicability.} What each method family requires of its target.
``Intervenes'' marks oracle-like $\mathrm{do}(u)$ access; ``hypothesis'' marks methods that
receive a supplied circuit hypothesis to test. The raw-VCS column distinguishes analyses that
read machine state directly from those that need a differentiable or layered model. A blank
means the requirement does not apply.}
\label{supp:tab:applic}
\footnotesize
\begin{tabular}{@{}p{3.0cm}cccccc@{}}
\toprule
Family & raw VCS & policy & gradients & layers/attn & intervenes & hypothesis \\
\midrule
A neuroscience & yes & no & no & no & A2 only & no \\
B gradient & yes & no & yes & no & no & no \\
B intervention & yes & no & no & no & yes & no \\
C causal & yes & no & no & yes & yes & yes \\
C gradient (attr.\ patch) & yes & no & yes & yes & approx. & yes \\
C probing & yes & no & no & yes & no & no \\
C / A dim.\ reduction & yes & no & no & yes & no & no \\
oracle (control) & yes & no & no & yes & exact & --- \\
\bottomrule
\end{tabular}
\end{table}

\subsection{T3 semantic-label coverage across the 64-game roster}\label{supp:t3coverage}

The tier-3 (T3) ground truth is externally supplied: it consists of semantic object and position
labels imported from AtariARI~\cite{anand2019unsupervised} and OCAtari~\cite{delfosse2023ocatari},
not derived by our oracle. This subsection documents where such labels \emph{exist}, so that the
breadth and position analyses can, in principle, scale beyond the six-game core of
Table~\ref{supp:tab:coverage}. It reports label \emph{availability and provenance} only. It does
not report any result on the wider battery. The larger sweep is in progress and has produced no
scores yet; nothing below should be read as a score.

The counts are computed, not estimated. We intersect our bit-exact roster with the games that
carry a label in the installed source of each project: the OCAtari object-property tables and the
AtariARI \texttt{atari\_dict}. OCAtari is preferred for render-aligned position labels because its
RAM mode applies the render-time offsets that correct the raw-RAM misalignment~\cite{delfosse2023ocatari};
AtariARI labels are cross-checkable but do not carry those offsets. Table~\ref{supp:tab:t3coverage}
summarizes the intersection. Of the 64 bit-exact games, 54 carry an OCAtari label, which is the
maximal labeled subset: every one of the 22 AtariARI-labeled games is also in the OCAtari 54, so
the AtariARI-only count is $0$ and OCAtari alone fixes the ceiling. The 22 games covered by both
sources are the richest and the only cross-checkable ones. Ten games carry no T3 label from either
source and are excluded from any label-dependent scoring. In percent of the roster, OCAtari labels
84\%, both sources cover 34\%, and 16\% have no label at all.

\begin{table}[t]
\centering
\caption{\textbf{T3 semantic-label coverage.} Availability of external position/object labels
across the 64-game bit-exact roster, computed against the installed OCAtari object tables and the
AtariARI \texttt{atari\_dict}. This is a provenance record of which games \emph{could} enter a
label-dependent analysis; it reports no result on any of them. Every AtariARI-labeled game is also
OCAtari-labeled, so OCAtari alone defines the maximal labeled set of 54.}
\label{supp:tab:t3coverage}
\footnotesize
\setlength{\tabcolsep}{5pt}
\begin{tabular}{@{}p{6.4cm}rp{4.2cm}@{}}
\toprule
Set & Count & Note \\
\midrule
Our bit-exact roster & 64 & all Paper-1 games \\
OCAtari-labeled (offset-corrected) & 54 & largest labeled subset \\
AtariARI-labeled & 22 & all a subset of the OCAtari 54 \\
Covered by both sources & 22 & richest (cross-checkable) \\
AtariARI-only & 0 & every AtariARI game is in OCAtari \\
No T3 label at all & 10 & excluded from label-dependent scoring \\
\bottomrule
\end{tabular}
\end{table}

For traceability we enumerate the sets. The 54 OCAtari-labeled games are:
\texttt{air\_raid}, \texttt{alien}, \texttt{amidar}, \texttt{assault}, \texttt{asterix},
\texttt{asteroids}, \texttt{atlantis}, \texttt{bank\_heist}, \texttt{battle\_zone},
\texttt{beam\_rider}, \texttt{berzerk}, \texttt{bowling}, \texttt{boxing}, \texttt{breakout},
\texttt{carnival}, \texttt{centipede}, \texttt{chopper\_command}, \texttt{crazy\_climber},
\texttt{demon\_attack}, \texttt{double\_dunk}, \texttt{enduro}, \texttt{fishing\_derby},
\texttt{freeway}, \texttt{frostbite}, \texttt{gopher}, \texttt{hero}, \texttt{ice\_hockey},
\texttt{jamesbond}, \texttt{kangaroo}, \texttt{krull}, \texttt{kung\_fu\_master},
\texttt{montezuma\_revenge}, \texttt{ms\_pacman}, \texttt{name\_this\_game}, \texttt{pacman},
\texttt{phoenix}, \texttt{pitfall}, \texttt{pong}, \texttt{pooyan}, \texttt{private\_eye},
\texttt{qbert}, \texttt{riverraid}, \texttt{road\_runner}, \texttt{robotank}, \texttt{seaquest},
\texttt{skiing}, \texttt{space\_invaders}, \texttt{star\_gunner}, \texttt{tennis},
\texttt{time\_pilot}, \texttt{up\_n\_down}, \texttt{venture}, \texttt{video\_pinball}, and
\texttt{yars\_revenge}. The 22 games covered by both sources are:
\texttt{asteroids}, \texttt{berzerk}, \texttt{bowling}, \texttt{boxing}, \texttt{breakout},
\texttt{demon\_attack}, \texttt{freeway}, \texttt{frostbite}, \texttt{hero},
\texttt{montezuma\_revenge}, \texttt{ms\_pacman}, \texttt{pitfall}, \texttt{pong},
\texttt{private\_eye}, \texttt{qbert}, \texttt{riverraid}, \texttt{seaquest},
\texttt{space\_invaders}, \texttt{tennis}, \texttt{venture}, \texttt{video\_pinball}, and
\texttt{yars\_revenge}. The 10 games with no T3 label, dropped from any label-dependent scoring,
are: \texttt{defender}, \texttt{elevator\_action}, \texttt{gravitar}, \texttt{journey\_escape},
\texttt{solaris}, \texttt{surround}, \texttt{tutankham}, \texttt{videochess},
\texttt{wizard\_of\_wor}, and \texttt{zaxxon}.

\subsection{Scored battery: the 42-game applicable subset}\label{supp:scoredbattery}

The T3-coverage of \Sectionname\ref{supp:t3coverage} sets the outer breadth of the sweep,
but not every labeled game can measure faithfulness. The interpretability battery is scored on
42 games, not on all 54 T3-labeled games of Table~\ref{supp:tab:t3coverage}. We hold out 12
games for two methodological reasons, decided before scoring. This is a scope decision about
where the benchmark can measure anything, not a result. The excluded games keep their exact T1
and T2 ground truth on record; only their score cells are held out.

The first reason is a lack of causal structure at the shared analysis state. Eight games fail
the oracle cause-density gate: fewer than four candidate causes move the shared output above
the floor ($|\Delta y|>0.5$). A frame with almost no true causal signal cannot measure
faithfulness, and running methods there would only add noise to the leaderboard. These eight
are \texttt{crazy\_climber}, \texttt{enduro}, \texttt{pooyan}, \texttt{skiing},
\texttt{time\_pilot}, \texttt{asteroids}, \texttt{battle\_zone}, and \texttt{star\_gunner}. The
second reason is the absence of a moving target for a position method. Seven games carry only
static verified labels, such as score or life counters, so no sprite moves under intervention
and a position method has nothing to recover. These seven are \texttt{amidar}, \texttt{asterix},
\texttt{robotank}, \texttt{up\_n\_down}, \texttt{asteroids}, \texttt{battle\_zone}, and
\texttt{star\_gunner}. The two criteria overlap for exactly three games, namely
\texttt{asteroids}, \texttt{battle\_zone}, and \texttt{star\_gunner}, which fail both. Counting
each excluded game once, the union is 5 gate-only plus 4 static-only plus 3 that fail both,
which is 12, so the scored battery is $54-12=42$ games. Table~\ref{supp:tab:scoredbattery}
lists the 12 with their reason.

Each of the 42 scored games both clears the cause-density gate and carries a sprite that moves
under intervention. Hence every one of them contributes to the all-regime leaderboard and to
the position regime. The six core games of Table~\ref{supp:tab:coverage} are the subset of
these 42 that also carry the full tiered ground truth and the exact oracle.

\begin{table}[t]
\centering
\caption{\textbf{Games held out from the scored battery.} The 12 games excluded from the
42-game interpretability battery, with the methodological reason for each. ``Gate'' marks a
game that fails the oracle cause-density gate (fewer than four causes with $|\Delta y|>0.5$ at
the shared analysis state). ``Static'' marks a game whose verified labels are all static
(e.g.\ score or life counters), so no sprite moves under intervention for a position method to
recover. The two criteria overlap for three games marked ``gate + static'', so counting each
game once the union is 5 gate-only plus 4 static-only plus 3 both, which is 12, and the scored
battery is $54-12=42$. These games keep their exact T1/T2 ground truth
(\Sectionname\ref{supp:t3coverage}); only their score cells are held out.}
\label{supp:tab:scoredbattery}
\footnotesize
\setlength{\tabcolsep}{6pt}
\begin{tabular}{@{}p{2.6cm}p{1.6cm}p{6.0cm}@{}}
\toprule
Game & Reason & Why held out \\
\midrule
crazy\_climber & gate & too few causes clear the density floor \\
enduro & gate & too few causes clear the density floor \\
pooyan & gate & too few causes clear the density floor \\
skiing & gate & too few causes clear the density floor \\
time\_pilot & gate & too few causes clear the density floor \\
\midrule
amidar & static & verified labels static; no sprite moves under intervention \\
asterix & static & verified labels static; no sprite moves under intervention \\
robotank & static & verified labels static; no sprite moves under intervention \\
up\_n\_down & static & verified labels static; no sprite moves under intervention \\
\midrule
asteroids & gate + static & too few causes clear the density floor and no sprite moves under intervention \\
battle\_zone & gate + static & too few causes clear the density floor and no sprite moves under intervention \\
star\_gunner & gate + static & too few causes clear the density floor and no sprite moves under intervention \\
\bottomrule
\end{tabular}
\end{table}
