\section{What this paper does not claim}\label{supp:notclaimed}

The benchmark is sharp because its scope is narrow. Stating the boundaries keeps the result
from being read wider than the evidence allows. We do not claim to recover a program's
semantics from scratch. The study measures whether an explanation names the true causes and
regenerates the behavior on a machine whose answer is already known. It does not show that any
method discovers the meaning of a routine without that ground truth in hand. A faithful method
recovers the causal structure but not the meaning of the variables, and this remaining gap is
the point rather than a result to be explained away.

We do not evaluate learned agents. The subject throughout is the bit-exact VCS itself. The
candidate causes are its registers, RAM cells, and program variables, not the weights of a
trained policy. The neural-network parallels drawn in the discussion and in
Fig.~\ref{fig:representativeness} are documented analogues from the literature, not claims proven on
this benchmark. We label them as such, so that a reader does not take a mapped failure mode
for a demonstrated one.

We do not prove that neural-network interpretability is impossible. The VCS gap is a
necessary-condition stress test. A method that fails to recover the causal structure of a
fully known machine cannot be trusted to recover it on an unknown one. Yet, a method that
succeeds here is not thereby certified on networks. The ``floor'' framing is an
evidence-transfer argument, not a proven lower bound on neural-network interpretability.

We do not measure human plausibility. The plausibility axis is a documented method-tradition
proxy, computed from a fixed rubric rather than collected from raters, and we report it as a
proxy throughout. Its job is to expose the danger zone where a convincing explanation is
unfaithful, not to stand in for a human-subjects study. We also do not redistribute the ROMs.
The substrate is validated against an external reference, and the records are reproducible from
committed artifacts without the ROMs. The ROM provenance is given as hashes
(\Sectionname\ref{supp:romhashes}), so a third party supplies their own legally obtained copies.

% The two full-resolution figures (representativeness map, failure taxonomy) are not
% reproduced here: they duplicated the main-text Fig.~\ref{fig:representativeness} and
% Fig.~\ref{fig:taxonomy}, whose captions carry the analogue citations and provenance. The
% Supplement refers to those main-text figures directly.
