\section{ROM provenance: SHA-256 hashes}\label{supp:romhashes}

We do not redistribute the Atari~2600 game cartridges. They are third-party ROMs
acquired through the \texttt{AutoROM} retrieval procedure, following the practice of the
underlying emulator port~\cite{maier2025vcs}. To let a third party verify that they hold the
exact binaries we ran against, we report a SHA-256 hash for each ROM in the bit-exact
roster. The Data-availability statement points here.

Table~\ref{supp:tab:romhashes} lists all 64 games of the Paper-1 bit-exact roster, so the
hash record covers not only the six-game core that carries the full method battery
(Table~\ref{supp:tab:coverage}) but also the wider 54-game labeled breadth set
(Table~\ref{supp:tab:t3coverage}) reserved for the larger sweep. A dagger ($\dagger$) marks the
six core games scored in this Article. Each hash is computed over the raw cartridge image
with \texttt{shasum -a 256}; reproducing it requires only the retrieved binary and a standard
SHA-256 tool. The hashes fingerprint the exact inputs; they are provenance, not a result.

% \hash: typeset a 64-hex-character SHA-256 digest in a narrow column, inserting
% zero-width discretionary breaks between characters so the string wraps cleanly.
% Kept local to this S-file so it stays self-contained (mirrors \jkey in supplement.tex).
\makeatletter
\def\hash#1{{\ttfamily\footnotesize\@hashaux#1\relax}}
\def\@hashaux#1{\ifx\relax#1\else#1\discretionary{}{}{}\expandafter\@hashaux\fi}
\makeatother

\begin{longtable}{@{}p{3.0cm}p{9.0cm}@{}}
\caption{\textbf{SHA-256 hashes of the 64 bit-exact ROMs.} One row per game of the
Paper-1 roster. A dagger ($\dagger$) marks the six core games scored in this Article; the
remaining rows document the wider substrate (including the 54-game labeled breadth set) for
which the ROM binaries are pinned but no interpretability score is reported here. Hashes are
computed with \texttt{shasum -a 256} over the retrieved cartridge image and are a provenance
record only.}
\label{supp:tab:romhashes} \\
\toprule
Game & SHA-256 \\
\midrule
\endfirsthead
\multicolumn{2}{@{}l}{\footnotesize Table~\ref{supp:tab:romhashes} (continued)}\\
\toprule
Game & SHA-256 \\
\midrule
\endhead
\bottomrule
\endfoot
\input{supplement/rom_hash_rows}
\end{longtable}
